---
title: "Getting Started with srlars"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with srlars}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

## Introduction

`srlars` fits the Fast and Scalable Cellwise-Robust Ensemble (FSCRE) algorithm: a
competitive ensemble of `n_models` sparse sub-models, built on a cellwise-robust
foundation (Detect Deviating Cells imputation and wrapping-based robust correlations)
so that variable selection and prediction remain reliable when individual *cells* of
the predictor matrix -- not necessarily whole observations -- are contaminated.

This vignette walks through a single, fast-running example: simulating a small
contaminated dataset, fitting `srlars()`, and inspecting its two controls over how
much the sub-models may share variables (`max_share`) and how small they may end up
(`n_min`). It ends with a short note on `cv.srlars()`, which chooses `max_share`
automatically.

```{r setup}
library(srlars)
library(mvnfast)
library(cellWise)
```

## Simulating a small, cellwise-contaminated dataset

We simulate `p = 30` predictors in two correlated blocks of 3 active variables each
(6 active predictors total), and contaminate a fraction of the cells of the training
predictors with correlation outliers -- values that distort the multivariate
dependence structure without necessarily standing out in any single variable. The
test set is left clean, as would be the case for genuinely new data.

```{r sim-setup}
set.seed(0)

n <- 60          # training observations
m <- 500         # test observations
p <- 30          # predictors
p.active <- 6    # truly active predictors
group.size <- 3
rho.within <- 0.7
rho.between <- 0.1
snr <- 3
contamination.prop <- 0.15
gamma <- 3       # contamination strength
n_models <- 4

# Block-correlation structure among the active predictors
sigma.mat <- matrix(0, p, p)
sigma.mat[1:p.active, 1:p.active] <- rho.between
for (g in 0:(p.active / group.size - 1)) {
  idx <- (g * group.size + 1):(g * group.size + group.size)
  sigma.mat[idx, idx] <- rho.within
}
diag(sigma.mat) <- 1

true.beta <- c(runif(p.active, 0, 5) * (-1) ^ rbinom(p.active, 1, 0.7),
              rep(0, p - p.active))
sigma <- as.numeric(sqrt(t(true.beta) %*% sigma.mat %*% true.beta) / sqrt(snr))
```

```{r sim-contam}
contam_correlation <- function(X, prop, sigma_mat, gamma) {
  n <- nrow(X); p <- ncol(X)
  idx <- sample.int(n * p, size = round(n * p * prop))
  rows <- ((idx - 1) %% n) + 1
  cols <- ((idx - 1) %/% n) + 1

  for (i in 1:n) {
    J <- cols[rows == i]
    if (length(J) == 0) next
    if (length(J) == 1) { X[i, J] <- gamma * 3; next }
    SigmaJ <- sigma_mat[J, J, drop = FALSE]
    ev <- eigen(SigmaJ, symmetric = TRUE)
    vmin <- ev$vectors[, ncol(ev$vectors)]
    denom <- mahalanobis(t(vmin), center = rep(0, length(J)), cov = SigmaJ)
    X[i, J] <- gamma * sqrt(length(J)) * (vmin / sqrt(denom))
  }
  X
}

# Training data (contaminated predictors)
x_train <- mvnfast::rmvn(n, mu = rep(0, p), sigma = sigma.mat)
y_train <- as.numeric(x_train %*% true.beta + rnorm(n, 0, sigma))
x_train <- contam_correlation(x_train, contamination.prop, sigma.mat, gamma)
colnames(x_train) <- paste0("V", 1:p)

# Test data (clean)
x_test <- mvnfast::rmvn(m, mu = rep(0, p), sigma = sigma.mat)
y_test <- as.numeric(x_test %*% true.beta + rnorm(m, 0, sigma))
colnames(x_test) <- colnames(x_train)
```

A small helper to score selection (precision/recall against the known active set)
and prediction (MSPE, scaled by the noise variance) is all we need throughout:

```{r metrics-helper}
get_metrics <- function(fit) {
  coefs <- as.numeric(coef(fit))[-1]
  sel <- which(coefs != 0)
  truth <- which(true.beta != 0)

  precision <- length(intersect(sel, truth)) / max(length(sel), 1)
  recall <- length(intersect(sel, truth)) / length(truth)

  preds <- as.numeric(predict(fit, x_test))
  mspe <- mean((y_test - preds)^2) / sigma^2

  c(Precision = precision, Recall = recall, MSPE = mspe,
   Mean.active.size = mean(vapply(fit$active.sets, length, integer(1))))
}
```

## Fitting srlars()

By default, `max_share = 1`: the `n_models` sub-models are fully disjoint -- no
variable can be selected by more than one of them.

```{r fit-default}
fit_default <- srlars(x_train, y_train,
                      n_models = n_models,
                      tolerance = 1e-4,
                      x_preprocess = "ddc",
                      y_preprocess = "wrap",
                      cor_estimator = "wrap",
                      cv_preprocess = "global",
                      cv_fit = "huber",
                      cv_loss = "huber",
                      cv_folds = 5,
                      compute_coef = TRUE)

fit_default$active.sets
get_metrics(fit_default)
```

`coef()` and `predict()` average across the ensemble's sub-models:

```{r coef-predict}
coef(fit_default)[1:5] # intercept + first few slopes
predict(fit_default, x_test[1:5, ])
```

## Controlling how much sub-models can share: `max_share`

`max_share` (1 to `n_models`) relaxes disjointness: a variable may be selected by up
to `max_share` sub-models. At the other extreme, `max_share = n_models` places no
restriction on sharing at all -- sub-models are then free to converge on the same
variables. Refitting on the same data:

```{r fit-shared}
fit_shared <- srlars(x_train, y_train,
                     n_models = n_models,
                     max_share = n_models,
                     tolerance = 1e-4,
                     x_preprocess = "ddc",
                     y_preprocess = "wrap",
                     cor_estimator = "wrap",
                     cv_preprocess = "global",
                     cv_fit = "huber",
                     cv_loss = "huber",
                     cv_folds = 5,
                     compute_coef = TRUE)

rbind(`max_share = 1 (default)` = get_metrics(fit_default),
     `max_share = n_models`    = get_metrics(fit_shared))
```

Whether sharing helps or hurts is dataset-dependent: more sharing means fewer
independent sub-models exploring different variables (which can hurt recall in a
high-dimensional, many-active-variable setting like this one), but it also lets the
ensemble refit the same, well-supported variables multiple times when a dataset does
not need broad exploration. `cv.srlars()` (see below) picks `max_share` for you by
cross-validation rather than requiring this kind of manual comparison.

## Guaranteeing a minimum sub-model size: `n_min`

The selection loop can stop early -- across the *entire* ensemble at once -- as soon
as no sub-model's next candidate variable shows a sufficient cross-validated
improvement. On some datasets this can leave sub-models quite small. `n_min` sets a
floor: sub-models below it keep receiving their best available variable even if it
doesn't clear the usual improvement bar (though it can never violate the
`max_share` diversity restrictions above).

```{r fit-nmin}
fit_floor <- srlars(x_train, y_train,
                    n_models = n_models,
                    n_min = 5,
                    tolerance = 1e-4,
                    x_preprocess = "ddc",
                    y_preprocess = "wrap",
                    cor_estimator = "wrap",
                    cv_preprocess = "global",
                    cv_fit = "huber",
                    cv_loss = "huber",
                    cv_folds = 5,
                    compute_coef = TRUE)

rbind(`n_min = NULL (default)` = get_metrics(fit_default),
     `n_min = 5`               = get_metrics(fit_floor))
```

## Choosing `max_share` automatically: `cv.srlars()`

Rather than comparing `max_share` values by hand as above, `cv.srlars()` chooses it
by an outer cross-validation loop scored on held-out *ensemble* prediction error, and
returns the refit at the cross-validated optimum. The expensive cellwise-robust
preprocessing stage is computed once per outer fold and reused across every candidate
value, so this is efficient relative to a naive grid search. The object it returns
is classed so that `coef()` and `predict()` work on it exactly as they do above --
no new methods to learn. See `?cv.srlars` for details.
